Appendix A : Dipole-dipole interaction between two spins in solid

### 2022-08-22 Manuscrit Thèse/chapter 2/Chapter 2.tex
taratata

\documentclass[a4paper,11pt]{report}
\usepackage[]{amsmath}
\usepackage[]{physics} % \bra, \ket etc
\usepackage{graphicx} %Pour les figures je crois
\usepackage{hyperref}
\usepackage[
    backend=biber, 
    natbib=true,
    style=numeric,
    sorting=none, %Pour faire apparaitre les refs dans l'ordre
    hyperref=true
]{biblatex} %Imports biblatex package
\addbibresource{Bib_ch2.bib} %Import the bibliography file

@@

\title{}
\begin{document}
\chapter{Detection of dark spins via cross-relaxations}
In this chapter, we will see how the interplay between the spin and the optical properties of the NV centers can be transferred to other spins in the crystal by coupling them to NV centers. More specifically, I will discuss the first optical detection of two spin species, VH$^-$ and War1, in a CVD diamond by the mean of cross-relaxations with NV centers.
\section{Flip-Flops, double flips and cross-relaxation}
\label{Sec_CR}

@@

Fig. \ref{CR_shema} illustrates CR between a polarized NV center and an unpolarized second spin. When the two considered spin transitions become resonant, in this case by tuning a magnetic field, polarization from the NV center will be transferred to the second spin, meaning in this case that the NV center will end up less polarized and the second spin more polarized.

@@

We refer by dark spins to spin defects which do not interact with light, which in the case of diamond defects consist of every known spin defect bar the NV center and the more recently found ST1 center \citep{lee2013readout, john2017bright}.

@@

The VH$^-$ defect, prior to our observations, had only been observed \textit{via} electron paramagnetic resonance (EPR) in CVD grown diamond. It is formed by a vacancy with a hydrogen atom bonded to one of the four dangling carbon bond, with an extra electron trapped in the vacancy. This defect has a similar electronic structure to the NV$^-$ which results in an electronic spin-1 with a similar zero field splitting parameter (see Table [REF]). Although the presence of hydrogen in this defect was confirmed by isotopic study, it has been debated whether this defect was VH$^-$ or V$_2$H$^-$ due to a conflict with \textit{ab initio} calculations \citep{shaw2005importance}. The defect is therefore sometimes referred as V$_n$H$^-$. In this manuscript I will employ the original name VH$^-$.

@@

There can be several advantages to use relaxometry protocols, which measures a change in the NV $T_1$, over the more common magnetometry protocols that rely on a change in the NV $T_2$ or $T_2^*$: the first and most obvious one is that relaxometry does not need to use an external microwave field. This however comes at the cost of a greater precision needed on the external magnetic field. Relaxometry also allows AC-magnetic field sensing in the GHz regime, where traditional microwave-based AC-megnetometry is limited to tens of MHz \citep{wang2022picotesla, alsid2022solid}. Finally, relaxometry measurement can be more sensitive than the T2-based one in some instances \citep{steinert2013magnetic}, which comes from the fact that changes in NV $T_1$ times are easier to detect than changes in $T_2$ or $T_2^*$ since $T_1$ is much longer than $T_2$ or $T_2^*$.

@@

Fig. \ref{121 vs 22 VH} shows the transition frequencies for the NV$^-$ and VH$^-$ ground state spin. Contrary to the P1 case, there is a co-resonance when $\mathbf{B} \parallel [100]$ for $B\approx 55\ \rm G$. When $\mathbf{B} \parallel [111]$, there are 6 co-resonance conditions between 30 and 130 G. A scan along the [111] in this case means that the peak signal is $\sim 6$ times weaker than in the [100] case, and spread over multiple features which might overlap between themselves or with other existing PL features.

@@

\begin{figure}[h]
\centering
\includegraphics[width=0.9\textwidth]{Figures/121_vs_22_13CNV}
\caption{a) Representation of the $^{13}$C-NV complex (right) compared to a "normal" NV center (left) b) and c) simulation of the transition frequencies of NV$^-$ centers and $^{13}$C-NV$^-$ complex for a magnetic field b) along the [100] axis and c) along the [111] axis}
\label{121 vs 22 13C-NV}
\end{figure}

@@

The $^{13}$C-NV$^-$ complex still behaves as a "normal" NV center, except with shifted transition lines. In particular its electronic spin is still polarized in the $\ket{0}$ state and the $\ket{0}$ state is still brighter than the $\ket{\pm 1}$ states. This is confirmed by the presence of sideband in ODMR spectra around the main NV lines \citep{simanovskaia2013sidebands}.

@@

Fig. \ref{121 vs 22 13C-NV}-b) and c) show the simulated transition frequencies for a $^{13}$C-NV$^-$ complex when B is scanned along the [100] or [111] axis, as well as the "normal" NV transitions and the co-resonance between those two. When $\mathbf{B} \parallel [100]$, there are 4 co-resonances between 18 and 25 G. When $\mathbf{B} \parallel [100]$ there are 18 of them between 10 and 60 G. This abundance of transitions for low ($<100\ \rm G$) magnetic field makes analyzing experimental data quickly intractable in the second case.

@@

Fig. \ref{CR VH exp}-a) shows the change in PL with respect to a magnetic field scanned along the diamond [100] axis. Similarly to fig. \ref{CR P1 exp}, we can see a slowly decaying envelope of the PL due to the transverse magnetic field (we don't see the plateau here since the magnetic field sops at 150 G). We can also see three deviations to the smooth decreasing curve at $\sim$ 20 G, 55 G and 120 G.

In order to better discern the three features, I have fitted the envelope by using a polynomial fit of order 4 on the smooth part of the data. I previously tried to use a fit based on the expected PL rate of the NV centers, by using the model in \citep{tetienne2012magnetic}, but the results were not satisfying enough.

Fig. \ref{CR VH exp}-b) shows the subtraction of the PL data by the envelope fit. We can clearly see three dips in PL around the values 20 G, 56 G and 122 G. I then computed the expected value of the magnetic field for the expected CR line thanks to the simulations shown in sec. \ref{sec simu}. For the VH$^-$ and War1 lines, I took the ZFS $D$ values from \citep{cruddace2007magnetic} and included the error bars. For the $^{13}$C-NV$^-$ lines, the error bars given in \citep{simanovskaia2013sidebands} are much smaller and I only plotted the expected value. We can see that the dips in PL match relatively well the predicted CR lines of $^{13}$C-NV$^-$, VH$^-$ and War1 respectively.

@@

The data shown here was acquired over 24 h. It should be noted however that a single-photon counting APD was used here, which considerably worsens the sensitivity of the measurement as discussed in [REF]. Using a photodiode and modulation of the magnetic field, I estimate that I could get similar signal to noise ratio within a few minutes.

To calibrate the magnetic field, I added a microwave field at a  fixed frequency $\nu > 2870\ \rm MHz$, and measured the EM voltage at which the PL dropped due to the $\ket{+1}$ state of the NV centers becoming resonant with the microwave tone. I could then use the values in Fig. \ref{121 vs 22 NV} to find the corresponding magnetic field. I varied the frequency in the range  $[2870,3150]\ \rm MHz$ by step of 2 MHz to get a calibration of the magnetic field in the $[0,150]\ \rm G$ region. Considering the uncertainty in the field angle, the NV ZFS $D$ value and the width of the PL drop, I estimate the final precision of the calibration to be within $\pm 1\ \rm G$.

@@

The second reason comes from the nature of the sample used. We tried similar PL scans along the [100] axis on many samples with dense ensemble of NV centers. We found that the dip at 55 G was present on all $\sim 5$ CVD samples used, and none of the $\sim 20$ HPHT samples. The same is true of the 122 G line, except that it did appear on some HPHT samples, as will be discussed below. We know that the VH$^-$ defect is much more likely to appear in a CVD sample due to the large quantity of hydrogen used in the process, and while we don't know the chemical nature of the War1 defect, the only time it was observed through EPR was on a CVD diamond.

@@

For this two reasons, we attribute the 20 G line to CR between $^{13}$C-NV$^-$ and "normal" NV centers.

@@

@@

\begin{itemize}
\item Since the change in $T_1$ is proportional to the $T_2^*$ of the probed spins (eq. \ref{delta gamma 1}), we should compensate that by integrating over the entire feature, instead of simply looking at the maximum of the line.
\item We have to take into account the loss of contrast of the NV centers as the transverse magnetic field grows, as explained in sec. \ref{CR contrast}.
\item The $^{13}$C-NV$^-$ are not "dark spins", and its CR feature cannot be \textit{a priori} compared to the one from VH$^-$ and War1. We will see however in the next chapter that we can consider a fraction of the NV centers as dark spins, and that this fraction can be estimated. \citep{choi2017depolarization} found a ratio of $\sim 1/4$ dark NV centers for $\sim 3/4$ bright ones.
\end{itemize}

@@

I Find that the area under 0 for the VH$^-$ line is $\approx 1.4$ times bigger than that of the $^{13}$C-NV$^-$ line, from \ref{121 vs 22 contrast} I find that the expected CR contrast is $\approx 1.2$ times smaller at 55 G than at 20 G, and finally I chose the dark to bright NV ratio value from \citep{choi2017depolarization} at 0.25.

With an estimated NV concentration [NV] $\approx\ \rm 4.6 ppm$, giving us a concentration $[^{13}\rm{C-NV}^-]\approx \ \rm 130 ppb$, I then find $[\rm VH^-] \approx 58\ \rm ppb$. Similarly, I find $[\rm VH^-] \approx 68\ \rm ppb$. Given the large uncertainty on some parameters, in particular the dark to bright spin ratio, these estimates are probably only precise within an order of magnitude.

@@

Fig \ref{chauffage CVD}-a) shows the demodulated PL versus mganetic field amplitude along [100] on sample CVD-C, while \ref{chauffage CVD}-b) shows the same experiment on sample CVD-B. The fabrication and subsequent operations done on this two samples are detailed thoroughly in \citep{ngambou2022improving}, but the main difference between these two sample is that CVD-C was only annealed once after irradiation, for 2 hours at 800$^\circ$C, while CVD-B was annealed for for 2 hours at 800$^\circ$C and then for 1 hour at 1000$^\circ$C and 1200$^\circ$C.

@@

An unexpected side-study was done on HPHT samples. Indeed, the CVD samples used previously are grown by homoepithaxy on top of a diamond substrate. In this case, the substrate was a type 1B HPHT diamond, which was irradiated and annealed along the CVD layer. I Will name SBST-C and SBST-B the respective HPHT substrate of CVD-C and CVD-B.

@@

I first suspected that some of the War1 defects had been migrating from the CVD layer to the HPHT substrate with the high-temperature annealing, but not only is it very unlikely with the annealing time (1 hour) and substrate thickness (several mm), I also could not find a dependence on the War1 concentration with the depth in the substrate, whereas a diffusion would have given a higher concentration near the CVD layer. The War1 defects seem to have been created by the annealing.

These results suggest that, while some defects are removed by the high temperature annealing, such as the VH$^-$ and several unidentified defects in the HPHT substrate, some on the other hand seem to be created by the annealing process, such as the War1 or the one associated with the line at $B = 48\ \rm G$ in the HPHT substrate.

@@

To conclude this chapter, we have seen how NV centers can be used to detect dark spins via cross-relaxations. This detection can be way more sensitive than the usual EPR spectroscopy, and considerably cheaper to setup. It is however not as versatile and should be seen as a complimentary tool for diamonds with dense NV ensemble.

We have also seen the importance of the magnetic field direction, and the underused aspect of the [100] magnetic field scan, much less noisy for weak magnetic field.

@@

I estimated the concentration of VH$^-$ in CVD-pink to be $\sim 60\ \rm ppb$. This concentration is too high by about two orders of magnitude to isolate a single emitter with a diffraction-limited microscope. We know however that we can reduce the VH$^-$ concentration arbitrarily by applying high temperature annealing. We could therefore reduce the VH$^-$ concentration to be in the $0.1 \sim 1\ \rm ppb$ range and isolate a single VH$^-$ center by scanning the sample with a fixed $B_{[100]}=56 \rm G$ while monitoring the NV PL or $T_1$. It is however unknown to me whether this detection scheme is sensitive enough to detect single dark spins.

>>>>

\documentclass[a4paper,11pt]{report}
\usepackage[]{amsmath}
\usepackage[]{physics} % \bra, \ket etc
\usepackage{graphicx} %Pour les figures je crois
\usepackage{hyperref}
\usepackage[
    backend=biber, 
    natbib=true,
    style=numeric-comp,
    sorting=none, %Pour faire apparaitre les refs dans l'ordre
    hyperref=true
]{biblatex} %Imports biblatex package
\addbibresource{Bib_ch2.bib} %Import the bibliography file

@@

\title{}
\begin{document}
\chapter{Detection of dark spins via cross-relaxations}
In this chapter, we will see how the interplay between the optical and spin properties of the NV centers can be transferred to other spins in the crystal by coupling them to NV centers. More specifically, I will discuss the first optical detection of two spin species, VH$^-$ and War1, in a CVD diamond by the mean of cross-relaxations with NV centers.
\section{Flip-Flops, double flips and cross-relaxation}
\label{Sec_CR}

@@

Fig. \ref{CR_shema} illustrates CR between a polarized NV center and an unpolarized second spin. When the two considered spin transitions become resonant, in this case by tuning the magnetic field, polarization from the NV center will be transferred to the second spin, making the NV center less polarized and the second spin more polarized than when they are out of resonance.

@@

We refer by dark spins to spin defects which do not interact with light, which in the case of diamond defects consist of almost every known spin defect bar the NV center and the more recently found ST1 center \citep{lee2013readout, john2017bright}.

@@

The VH$^-$ defect, prior to our observations, had only been observed \textit{via} electron paramagnetic resonance (EPR) in CVD grown diamond. It is formed by a vacancy with a hydrogen atom bonded to one of the four dangling carbon bond, with an extra electron trapped in the vacancy. This defect has a similar electronic structure to the NV$^-$ which results in an electronic spin-1 with a similar zero field splitting parameter (see Table \ref{table VH et War1}). Although the presence of hydrogen in this defect was confirmed by isotopic study, it has been debated whether this defect was VH$^-$ or V$_2$H$^-$ due to a conflict with \textit{ab initio} calculations \citep{shaw2005importance}. The defect is therefore sometimes referred as V$_n$H$^-$. In this manuscript I will employ the original name VH$^-$.

@@

There can be several advantages to use relaxometry protocols, which measures a change in the NV $T_1$, over the more common magnetometry protocols that rely on a change in the NV $T_2$ or $T_2^*$: the first and most obvious one is that relaxometry does not need to use an external microwave field. This however comes at the cost of a greater precision needed on the external magnetic field. Relaxometry also allows AC-magnetic field sensing in the GHz regime \citep{wang2022picotesla, alsid2022solid}, while traditional microwave-based AC-megnetometry is limited to tens of MHz. Finally, relaxometry measurement can be more sensitive than the T2-based one in some instances \citep{steinert2013magnetic}, which comes from the fact that changes in NV $T_1$ times are easier to detect than changes in $T_2$ or $T_2^*$ since $T_1$ is much longer than $T_2$ or $T_2^*$.

@@

Fig. \ref{121 vs 22 VH} shows the transition frequencies for the NV$^-$ and VH$^-$ ground state spin. Contrary to the P1 case, there is a co-resonance when $\mathbf{B} \parallel [100]$ for $B\approx 55\ \rm G$. When $\mathbf{B} \parallel [111]$, there are 6 co-resonance conditions between 30 and 130 G. A scan along the [111] in this case means that the peak signal will be $\sim 6$ times weaker than in the [100] case, and spread over multiple lines which might overlap between themselves or with other existing PL features.

@@

\begin{figure}[h]
\centering
\includegraphics[width=1\textwidth]{Figures/121_vs_22_13CNV}
\caption{a) Representation of the $^{13}$C-NV complex (right) compared to a ``normal" NV center (left) b) and c) simulation of the transition frequencies of NV$^-$ centers and $^{13}$C-NV$^-$ complex for a magnetic field b) along the [100] axis and c) along the [111] axis}
\label{121 vs 22 13C-NV}
\end{figure}

@@

The $^{13}$C-NV$^-$ complex still behaves as a ``normal" NV center, except that its transition lines are shifted by the hyper-fin coupling. In particular its electronic spin is still polarized in the $\ket{0}$ state and the $\ket{0}$ state is still brighter than the $\ket{\pm 1}$ states. This is confirmed by the presence of sideband in ODMR spectra around the main NV lines \citep{simanovskaia2013sidebands}.

@@

Fig. \ref{121 vs 22 13C-NV}-b) and c) show the simulated transition frequencies for a $^{13}$C-NV$^-$ complex when B is scanned along the [100] or [111] axis, as well as the ``normal" NV transitions and the co-resonance between those two. When $\mathbf{B} \parallel [100]$, there are 4 co-resonances between 18 and 25 G. When $\mathbf{B} \parallel [100]$ there are 18 of them between 10 and 60 G. This abundance of transitions for low ($<100\ \rm G$) magnetic field makes analyzing experimental data quickly intractable in the second case.

@@

Fig. \ref{CR VH exp}-a) shows the change in PL with respect to a magnetic field scanned along the diamond [100] axis. Similarly to fig. \ref{CR P1 exp}, we can see a slowly decaying envelope of the PL due to the transverse magnetic field (we don't see the plateau here since the magnetic field stops at 150 G). We can also see three deviations to the smooth decreasing curve at $\sim$ 20 G, 55 G and 120 G.

In order to better discern the three features, we fitted the envelope by using a polynomial fit of order 4 on the smooth part of the data. We previously tried to use a fit based on the expected PL rate of the NV centers, by using the model in \citep{tetienne2012magnetic}, but the results were not satisfying.

Fig. \ref{CR VH exp}-b) shows the subtraction of the PL data by the envelope fit. We can clearly see three dips in PL around the values 20 G, 56 G and 122 G. We then computed the expected value of the magnetic field for the expected CR line thanks to the simulations shown in sec. \ref{sec simu}. For the VH$^-$ and War1 lines, we took the ZFS $D$ values from \citep{cruddace2007magnetic} and included the error bars. For the $^{13}$C-NV$^-$ lines, the error bars given in \citep{simanovskaia2013sidebands} are much smaller and we only plotted the expected value. We can see that the dips in PL match relatively well the predicted CR lines of $^{13}$C-NV$^-$, VH$^-$ and War1 respectively.

@@

The data shown here was acquired over 24 h. It should be noted however that a single-photon counting APD was used here, which considerably worsens the sensitivity of the measurement as discussed in [REF]. Using a photodiode and modulation of the magnetic field, we estimate that we could get similar signal to noise ratio within a few minutes.

To calibrate the magnetic field, we added a microwave field at a  fixed frequency $\nu > 2870\ \rm MHz$, and measured the EM voltage at which the PL dropped due to the $\ket{+1}$ state of the NV centers becoming resonant with the microwave tone. We could then use the values in Fig. \ref{121 vs 22 NV} to find the corresponding magnetic field. We varied the frequency in the range  $[2870,3150]\ \rm MHz$ by step of 2 MHz to get a calibration of the magnetic field in the $[0,150]\ \rm G$ region. Considering the uncertainty in the field angle, the NV ZFS $D$ value and the width of the PL drop, we estimate the final precision of the calibration to be within $\pm 1\ \rm G$.

@@

The second reason comes from the nature of the sample used. We tried similar PL scans along the [100] axis on many samples with dense ensemble of NV centers. We found that the dip at 55 G was present on all CVD samples ($\sim$ 5 samples) used, and none of the HPHT samples ($\sim$ 20 samples). The same is true of the 122 G line, except that it did appear on some HPHT samples, as will be discussed below. We know that the VH$^-$ defect is much more likely to appear in a CVD sample due to the large quantity of hydrogen used in the process, and while we don't know the chemical nature of the War1 defect, the only time it was observed through EPR was on a CVD diamond.

@@

For this two reasons, we attribute the 20 G line to CR between $^{13}$C-NV$^-$ and ``normal" NV centers.

@@

This procedure will be discussed in more details in the next chapter. We find here that the linewidth of VH$^-$ and War1 are respectively [REF] and [REF] MHz.

@@

\begin{itemize}
\item Since the change in $T_1$ is proportional to the $T_2^*$ of the probed spins (eq. \ref{delta gamma 1}), we should compensate that by integrating over the entire feature, instead of simply looking at the maximum of the line.
\item We have to take into account the loss of contrast of the NV centers as the transverse magnetic field grows, as explained in sec. \ref{CR contrast}.
\item The $^{13}$C-NV$^-$ are not ``dark spins", and its CR feature cannot be \textit{a priori} compared to the one from VH$^-$ and War1. We will see however in the next chapter that we can consider a fraction of the NV centers as dark spins, and that this fraction can be estimated. \citep{choi2017depolarization} found a ratio of $\sim 1/4$ dark NV centers for $\sim 3/4$ bright ones.
\end{itemize}

@@

For the VH$^-$ line we find that the area of the line is $\approx 1.4$ times bigger than that of the $^{13}$C-NV$^-$ one, from \ref{121 vs 22 contrast} we find that the expected CR contrast is $\approx 1.2$ times smaller at 55 G than at 20 G, and finally we chose the dark to bright NV ratio value from \citep{choi2017depolarization} at 0.25.

With an estimated NV concentration [NV] $\approx\ \rm 4.6 ppm$, giving us a concentration $[^{13}\rm{C-NV}^-]\approx \ \rm 130 ppb$, we then find $[\rm VH^-] \approx 58\ \rm ppb$. Similarly, we find $[\rm VH^-] \approx 68\ \rm ppb$. Given the large uncertainty on some parameters, in particular the dark to bright spin ratio, these estimates are probably only precise within an order of magnitude.

@@

Fig \ref{chauffage CVD}-a) shows the demodulated PL versus mganetic field amplitude along [100] on sample CVD-C, while \ref{chauffage CVD}-b) shows the same experiment on sample CVD-B. The fabrication and subsequent operations done on this two samples are detailed thoroughly in \citep{ngambou2022improving}, but the main difference between these two sample is that CVD-C was only annealed once after irradiation, for 2 hours at 800$^\circ$C, while CVD-B was annealed for 2 hours at 800$^\circ$C and then for 1 hour at 1000$^\circ$C and 1 hour at 1200$^\circ$C.

@@

An unexpected side-study was done on HPHT samples. Indeed, the CVD samples used previously are grown by homoepithaxy on top of a diamond substrate. In this case, the substrate was a type 1B HPHT diamond, which was irradiated and annealed along the CVD layer. We Will name SBST-C and SBST-B the respective HPHT substrate of CVD-C and CVD-B.

@@

We first suspected that some of the War1 defects had been migrating from the CVD layer to the HPHT substrate with the high-temperature annealing, but not only is it very unlikely with the annealing time (1 hour) and substrate thickness (several mm), We also could not find a dependence on the War1 concentration with the depth in the substrate, whereas a diffusion would have given a higher concentration near the CVD layer. The War1 defects seem to have been created by the annealing.

These results suggest that, while some defects are removed by the high temperature annealing, such as the VH$^-$ and several unidentified defects in the HPHT substrate, some seem to be created by the annealing process, such as the War1 or the one associated with the line at $B = 48\ \rm G$ in the HPHT substrate.

@@

To conclude this chapter, we have seen how NV centers can be used to detect dark spins via cross-relaxations. This detection can be way more sensitive than the usual EPR spectroscopy, and considerably cheaper to set up. It is however not as versatile and should be seen as a complimentary tool for diamonds with dense NV ensemble.

We have also seen the importance of the magnetic field direction, and the underused aspect of the [100] magnetic field scan, much less noisy than the [111] scans for weak magnetic field.

@@

We estimated the concentration of VH$^-$ in CVD-pink to be $\sim 60\ \rm ppb$. This concentration is too high by about two orders of magnitude to isolate a single emitter with a diffraction-limited microscope. We know however that we can reduce the VH$^-$ concentration arbitrarily by applying high temperature annealing. We could therefore reduce the VH$^-$ concentration to be in the $0.1 \sim 1\ \rm ppb$ range and optically isolate a single VH$^-$ center by scanning a confocal microscope with a fixed $B_{[100]}=56\ \rm G$ while monitoring the NV PL or $T_1$. It is however unclear to me whether this detection scheme would be sensitive enough to detect single dark spins.
